---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

# SQL R Tools

![](https://img.shields.io/badge/lifecycle-Stable-green.svg)

The aim of this package is to provide functionality to support working with 
MSSQL and MYSQL databases from R.

The package currently contains one R6 class object that is used to connect with
a given SQL database and provides several methods for interacting with the 
database, including:

* **close_connection**: close the connection to the server, note that other
methods will do this by default. 
* **get**: run SQL query on the server and return results to R environment. 
* **run**: run SQL query on the server without returning results to R
environment, e.g. if inserting data into another table/creating temp tables etc. 
* **table_exists**: check if a table exists in a database. 
* **upload**: upload data to a database, with options to batch upload. 
* **drop_table**: drop table in a database.
* **databases**: list databases in a server.
* **db_tables**: list tables in a database.
* **db_views**: list views in a database.
* **temp_table_name**: get the full name of a temporary table. 
* **object_fields**: get list of fields from a table/view.
* **order_object_fields**: lists the fields in a table/view and order by data
type, with varchar(max) and geometry fields at the end (i.e. useful for 
avoiding "Invalid Descriptor Index" errors, see examples below)
* **meta_data**: gives details on given list of tables/views, including data
types, indexes, completeness and proportion values that are unique. 

## Installation

This package is currently private and therefore requires a GitHub Personal 
Access Token (PAT) to be set up for an account that has access to the 
NottsHC GitHub organisation on GitHub. 

**Important**: your GitHub PAT should be stored in the local Git credential store
and **not** a local .Renviron file. Further to this, the package `keyring`
should be used to store credentials used in scripts rather than the 
Renviron file, see recommendations on this 
[here](https://solutions.posit.co/connections/db/best-practices/managing-credentials/).
Note that the `get_env_var()` function in this package should be used in place
of the `get_env_var()` function, which uses `keyring` in the background and 
prompts moving from Renviron to `keyring`where relevant. 

To create a GitHub PAT (if you don't have one already):

1. Go to <https://github.com/settings/tokens> which takes you to
your own personal GitHub account settings where you will need to
`Generate new token`. 
2. Give this a suitable name like `Repo access` and tick the `repo` group for 
full control of private repositories.
3. *Remember to copy the code that is generated as this cannot be viewed
again*

To set up GitHub PAT in the Git Credential Store & install package:

```{r, install, eval=FALSE}
# Install required packages
install.packages("usethis")
install.packages("gitcreds")
install.packages("remotes")
install.packages("pak")

# set up GitHub PAT. Note: if a GitHub PAT has already been saved, the following
# will give an option to replace the credentials. Run the function, select the
# relevant option and update the PAT with the one just copied from GitHub.
gitcreds::gitcreds_set()

# Check has been set properly, the below should return a list with details of:
# protocol, host, username and password. None of these should be NA, the
# password should have "<-- hidden -->".
gitcreds::gitcreds_get()

# If the above doesn't look correct:
#  1. Make sure you don't have GITHUB_PAT set in your Renviron file. If you do,
#     delete it, restart R, and check it's no longer in use by running
#     Sys.getenv("GITHUB_PAT"), this should return "".
#  2. Delete your GITHUB_PAT on GitHub at https://github.com/settings/tokens and
#     re-create the PAT using the instructions above (this may have failed if
#     you already had a GitHub PAT set up).

# If the output of gitcreds::gitcreds_get() still looks incorrect, contact the
# author of this package.

# install the package (see notes below if this fails)
pak::pkg_install("Notts-HC/SQLRtools")
```

If you have a `GITHUB_PAT` set up in the Renviron file **delete it**. The 
following can be used to check this: `get_env_var("GITHUB_PAT")`, it should
return "". 

Full details on this recommended approach can be found 
[here](https://usethis.r-lib.org/articles/git-credentials.html).

The above should install the package **but will fail if you don't have the 
correct version of RTools installed to work with your version of R**, i.e.
if you've updated to R v 4.4.1 but still only have RTools4 you'll get an
error around the version of Rtools or R not being able to locate the relevant
build tools. A request has been made to IT to update the version of RTools 
available in the software center. 

However,the following work around can be used to install with remotes by using the
GitHub PAT as an environment variable via git credentials:


```{r, install back up, eval = FALSE}
# the following can be used as a work around if pak fails to install the package:

# set the GITHUB_PAT environment vairbale (this will NOT be saved in Renviron):
Sys.setenv(GITHUB_PAT = gitcreds::gitcreds_get(use_cache = FALSE)$password)

# install using remotes
remotes::install_github("Notts-HC/SQLRtools")
```

## About

You are reading the doc about version: 0.0.1

This README has been compiled on the

```{r}
Sys.time()
```

Here are the tests results and package coverage:

```{r, error = TRUE}
devtools::check(quiet = TRUE)
```

```{r echo = FALSE}
unloadNamespace("SQLRtools")
```

```{r, error = TRUE}
covr::package_coverage()
```

## Using the package

As above, the package contains an R6 class object that acts as the connection
to the server. This means that once the initial sql_server class object is 
created there is no further need to provide connection details to connect to the
server. 

The below gives examples of connecting to a MSSQL server and MYSQL server:

```{r sql connection, eval=FALSE}
library(SQLRtools)


# set connection to MS SQL server
ms_sql_server <- sql_server$new(
  driver = "SQL Server",
  server = get_env_var("MSSQL_SERVER"),
  database = get_env_var("MSSQL_DATABASE")
)

# set connect to MySQL server
my_sql_server <- sql_server$new(
  driver = "MySQL ODBC 8.0 Unicode Driver",
  server = get_env_var("HOST_NAME"),
  database = get_env_var("MYSQL_DB"),
  port = get_env_var("MYSQL_PORT"),
  uid = get_env_var("MYSQL_USER"),
  pwd = get_env_var("MYSQL_PASSWORD")
)
```

The methods listed above can now be used with these connections to:

##### Upload & query data

```{r basic sql examples, eval=FALSE}
# create a basic temp table
my_data <- data.frame(
  a = c("a", "b", "c"),
  b = 1:3
)

# upload as a temporary table - note that generally close_conn should be TRUE
# (which is the default setting), but it needs to be FALSE here so the
# connection isn't shut after uploading the temporary table, as this would drop
# the temporary table straight away.
ms_sql_server$upload(
  data = my_data,
  table_name = "#SQLRtools_example",
  close_conn = FALSE
)

# get the data
sql_data <- ms_sql_server$get("SELECT *
                               FROM #SQLRtools_example")
```


##### Explore databases & their objects

```{r details of objects on server, eval=FALSE}
# get databases in server
ms_sql_dbs <- ms_sql_server$databases()

# get list of tables in a given database
my_sql_db_tables <- ms_sql_server$db_tables(database = ms_sql_dbs$name[20])

# get list of views in a given database
my_sql_db_views <- ms_sql_server$db_views(database = ms_sql_dbs$name[20])

# get meta data of table in given tables
my_sql_meta_data <- ms_sql_server$meta_data(
  database = ms_sql_dbs$name[20],
  objects = my_sql_db_views$view_name[1:5],
  details = FALSE
)

names(my_sql_meta_data)[1]
View(my_sql_meta_data[1][[1]])
```

##### Avoid the "Invalid Descriptor Index" issue

```{r invalid descriptor index example, eval=FALSE}
# create a basic temp table
my_data <- data.frame(
  a = c("a", "b", "c"),
  b = 1:3
)

# upload as a temporary table
ms_sql_server$upload(
  data = my_data,
  table_name = "#SQLRtools_example",
  close_conn = FALSE
)

# change the data type of column a to varchar(max)
ms_sql_server$run("ALTER TABLE #SQLRtools_example
                   ALTER COLUMN a varchar(max);",
  close_conn = FALSE
)

# try extracting the data
ms_sql_server$get("SELECT * FROM #SQLRtools_example",
  close_conn = FALSE
)

# use method order_object_fields to avoid issue

# get all the fields ordered by data type
table_fields <- ms_sql_server$order_object_fields(
  object = "#SQLRtools_example",
  close_conn = FALSE
)

# now extract all the data
ms_sql_server$get(glue("SELECT {table_fields}
                        FROM #SQLRtools_example"))

# note that the above table only has 2 fields. Where your trying to run
# select * on a table with a lot of fields and multiple varchar(max) or
# geometry feilds, the above makes life a lot easier (notwithstanding the fact
# it's good practice to avoid 'select *' in SQL where possible).
```
